---
title: "Project 1: Looking at Montgomery Country Employee Salaries of 2017"
author: "Lucy Murray"
date: "6/13/2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

For this project, I will be examining the annual salary information for Montgomery County, Maryland employees paid in calendar year 2017. It is available publicly here: https://data.montgomerycountymd.gov/Human-Resources/Employee-Salaries-2017/2qd6-mr43

### Importing Data and Storing it in "empsalaries2017"

First, I would like to import my data and store it in the variable "empsalaries2017" as this will make it easier to work with.

```{r}
getwd()
setwd("C:/Users/murrayl/Documents/Lucy Murray/Personal/DATA101/PROJECT 1")
empsalaries2017 <- read.csv("Employee_Salaries_-_2017.csv")

```

I will be using functions from the dplyr library. It can be installed with install.packages() if it is not already on the device.
```{r}
library(dplyr)
```

# The Data at a Glance

### Confirming the Class and Dimensions

I assume the imported data will be a data frame, as it contains data of different data types. I can confirm this by utilizing the class() function.

```{r}
#Viewing the class of the data set
class(empsalaries2017)

```

To ensure the data was imported correctly, I can confirm that the expected 9,398 rows and 12 columns imported correctly with the dim() function.

```{r}
#Viewing the dimensions of the data set
dim(empsalaries2017)

```

### Observing the Structure, Head, and Tail

For more detailed information, I can call the str() function, which provides the internal structure of my data set.

```{r}
#Structure of my data
str(empsalaries2017)

```

The 9,398 observations make this a huge data set! It might be impractical to call the entire data set.The heads() and tails() functions is perfect, as I can choose to observe just the first and last 10 rows.

```{r}
#First 10 rows
head(empsalaries2017, 10)

#Last 10 rows
tail(empsalaries2017, 10)
```

# Removing Missing Values Values
I noticed that many of the observations have missing values. I can remove all of the missing values and store it in the "complete_empsalaries2017" variable using the na.omit() function.
```{r}
#Remove observations with missing values
complete_empsalaries2017 <- na.omit(empsalaries2017)
```

I can now observe the structure of the new variable to see how many observations were removed. 
```{r}
#Checking structure of complete_empsalaries2017
str(complete_empsalaries2017)
```

The original data frame contained 9,398 observations and the new one contains 3,877 observations. The difference between these two indicates that 5,521 observations were removed!

# Selecting Columns
The select() function is useful for choosing which columns I would like to view in the data set. I can select columns by their names and indices. In the first example, I will select the columns with indices 1 through 5. In the second example, I will select columns with the names Full.Name, Gender and Current.Annual.Salary. I will store these results in new variables called "basics" and "basics2", respectively. For both, I will print out the head to ensure that my function worked correctly.
```{r}
#Selecting columns by indices 1 through 5
basics <- select(complete_empsalaries2017, 1:5)
head(basics)
#Selecting columns with 
basics2 <- select(complete_empsalaries2017, Full.Name, Gender, Current.Annual.Salary)
head(basics2)
```

# Filter
The filter() function works similarly to select. I can filter by a column's attributes. In this example, I will filter so that I only see observations that contain a value of greater than 80,000 under the Current.Annual.Salary column.
```{r}
basics3 <- filter(complete_empsalaries2017, Current.Annual.Salary>80000)
head(basics3)
```

# Reordering the data
If I am interested in ordering the dataframe, I can simply call index the data by treating it as a list. For both examples, I will call rows 1:5. I can call the columns by index number or name. In the first example, I will call only the Full.Name, Gender, and Current.Annual Salary columns, and I will reorder them by their indices. In the second example, I will reorder the columns by name.
```{r}
complete_empsalaries2017[1:5,c(1,3,2)]
complete_empsalaries2017[1:5,c("Full.Name","Assignment.Category","Current.Annual.Salary")]
```

# Subsetting

### Looking at Male and Female Employee Salaries Separately

I am interested in looking at the salaries of males compared to females. I can do so utilizing the subset() function and storing the observations with gender male in the new "male_emp" variable and observations with gender female in the new "female_emp" variable, respectively. I use head() the way I did before, to confirm that the subset function worked correctly.
```{r}
#Creating the variable for male employees
male_emp <- subset(complete_empsalaries2017, Gender=="M")

#Checking the head of male_emp
head(male_emp)

#Creating the variable for female employees
female_emp <- subset(complete_empsalaries2017, Gender=="F")

#Checking the head of female_emp
head(female_emp)
```

Let's make some quick comparisons! I can check out mininum, maximum, median, mean, 25th, and 75th quartiles utilizing the summary() function of the current annual salary for both male and female employees.

```{r}
#Observing summary statistics for male employees
summary(male_emp$Current.Annual.Salary)

#Observing summary statistics for female employees
summary(female_emp$Current.Annual.Salary)
```


### Looking at Full-time and Part-time Employee Salaries separately

I am interested in looking at the salaries of full-time compared to part-time employees. As before, I can do so utilizing the subset() function and storing the observations with full-time status in the new "full_emp" variable and observations with part-time status in the new "part_emp" variable, respectively. I will use head() the way I did before, to confirm that the subset function worked correctly.
```{r}
#Creating the variable for Full-time Employees
full_emp <- subset(complete_empsalaries2017, Assignment.Category=="Fulltime-Regular")

#Checking the head of male_emp
head(full_emp)

#Creating the variable for Part-time Employees
part_emp <- subset(complete_empsalaries2017, Assignment.Category=="Parttime-Regular")

#Checking the head of Part-time Employees
head(part_emp)
```

As I did for male and female employees, I will now check out mininum, maximum, median, mean, 25th, and 75th quartiles utilizing the summary() function of the current annual salary for both full-time and part-time employees.

```{r}
#Observing summary statistics for full-time employees
summary(full_emp$Current.Annual.Salary)

#Observing summary statistics for part-time employees
summary(part_emp$Current.Annual.Salary)
```